---
title: "A1_ChrisEscobedo"
author: "Chris"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document  
---

```{r, include = F}
knitr::opts_chunk$set(echo = F, warnings = F, message = F)
```

```{r, warning = F}
library(tidyverse)
library(mapview)
library(tigris)
library(sf)
library(leaflet)
library(esri2sf)
library(tidycensus)
library(censusapi)

Sys.setenv(CENSUS_KEY="c7ce771b506b13ca6875d12a179606f8087fa381")

census_api_key("c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")

acs_vars_2019_5yr <-
  listCensusMetadata(
    name = "2019/acs/acs5/subject",
    type = "variables"
  )
```

Intro

### Housing Burden

For the purpose of this assesment, housing burden has been defined as using by a threshold of 30% of one's household income. Those who spend more than 30% of income on housing are considered to be housing burdened, while those spend less are considered to not be housing burdened. Although this is an arbitrarily chosen cutoff, it is useful to use to get a quick understanding of those who may struggle to pay for housing in the area.

That being said, it is important to note there may be times when this makes the findings less accurate. For extremely high income earners, using more than 30% on housing may not matter to their living situation if they have plenty of disposable income leftover. Furthermore, some families that use 29% of their income on housing may be struggling financially, but will not be counted in this measure of housing burden because of the arbitrary cutoff. 


```{r, warning = F}
neighboring_counties <- 
  counties("CA", cb = T, progress_bar = F) %>% 
  filter(NAME %in% c("San Mateo","Santa Clara"))

ca_places <- 
  places("CA", cb = T, progress_bar = FALSE) 

neighboring_places <- 
  ca_places %>% 
  st_centroid() %>% 
  .[neighboring_counties, ] %>% 
  st_drop_geometry() %>% 
  left_join(ca_places %>% select(GEOID)) %>% 
  st_as_sf() 

EPA_boundary <- 
  places("CA", cb = T, progress_bar = FALSE) %>% 
  filter(NAME == "East Palo Alto")

housing_burden_neighboring_counties <- getCensus(
  name = "acs/acs5/subject",
  vintage = 2019,
  region = "place:*",
  regionin = "state:06",
  vars = paste0("S2503_C0", c(
  "3_025E","3_026E", "3_027E", "3_028E", "3_029E","3_030E", "3_031E", "3_032E", "3_033E", "3_034E", "3_035E", "3_036E", "3_037E", "3_038E", "3_039E", "3_040E", "3_041E", "3_042E", "3_043E", "3_044E", "3_045E", "3_046E", "5_025E","5_026E", "5_027E", "5_028E", "5_029E","5_030E", "5_031E", "5_032E", "5_033E", "5_034E", "5_035E", "5_036E", "5_037E", "5_038E", "5_039E", "5_040E", "5_041E", "5_042E", "5_043E", "5_044E", "5_045E", "5_046E")
  )
) %>% 
  filter(place %in% neighboring_places$PLACEFP) %>%
  select(!c(state,place) & !ends_with(c("EA","MA","M"))) %>%
  pivot_longer(
    ends_with("E"),
    names_to = "name",
    values_to = "estimate"
  ) %>%
  left_join(
    acs_vars_2019_5yr %>% 
      select(name, label)
    ) %>% 
  select(-name) %>% 
  separate(
    label,
    into = c(NA,"occupation",NA, NA,"income", "burden"),
    sep = "!!"
  ) %>% 
  filter(!is.na(burden)) %>% 
  mutate(
    burden = if_else(burden == "20 to 29 percent" | burden == "Less than 20 percent", 
              "less than 30 percent", 
              burden)
  ) %>% 
  group_by(income, occupation, burden) %>% 
  summarize(estimate = sum(estimate))

housing_burden_epa <- getCensus(
  name = "acs/acs5/subject",
  vintage = 2019,
  region = "place:20956",
  regionin = "state:06",
  vars = paste0("S2503_C0", c(
  "3_025E","3_026E", "3_027E", "3_028E", "3_029E","3_030E", "3_031E", "3_032E", "3_033E", "3_034E", "3_035E", "3_036E", "3_037E", "3_038E", "3_039E", "3_040E", "3_041E", "3_042E", "3_043E", "3_044E", "3_045E", "3_046E", "5_025E","5_026E", "5_027E", "5_028E", "5_029E","5_030E", "5_031E", "5_032E", "5_033E", "5_034E", "5_035E", "5_036E", "5_037E", "5_038E", "5_039E", "5_040E", "5_041E", "5_042E", "5_043E", "5_044E", "5_045E", "5_046E")
  )
) %>% 
  filter(place %in% neighboring_places$PLACEFP) %>%
  select(!c(state,place) & !ends_with(c("EA","MA","M"))) %>%
  pivot_longer(
    ends_with("E"),
    names_to = "name",
    values_to = "estimate"
  ) %>%
  left_join(
    acs_vars_2019_5yr %>% 
      select(name, label)
    ) %>% 
  select(-name) %>% 
  separate(
    label,
    into = c(NA,"occupation",NA, NA,"income", "burden"),
    sep = "!!"
  ) %>% 
  filter(!is.na(burden)) %>% 
  mutate(
    burden = if_else(burden == "20 to 29 percent" | burden == "Less than 20 percent", 
              "less than 30 percent", 
              burden),
    occupation = if_else(occupation == "Owner-occupied housing units", 
              "Owner", 
              "Renter"),
  ) %>% 
  group_by(income, occupation, burden) %>% 
  summarize(estimate = sum(estimate))

housing_burden_epa_reordered <- housing_burden_epa %>% 
  filter(income == "Less than $20,000") %>% 
  rbind(housing_burden_epa) %>% 
  .[-c(21, 22, 23, 24), ]



```

### Housing Burden in East Palo Alto 

Below are two plots that represent general housing burden in East Palo Alto. The plots are separated by occupational status as either a "renter" or "owner" of the house. Since the OPA is interested in assisting the renter-occupied families who are renting housing in building wealth and preventing their displacement, it is helpful to gain this general understanding of the renting and housing burden landscape.

```{r, warning = F}
burden_plot_epa <- housing_burden_epa_reordered %>% 
  ggplot() +
  geom_bar(
    aes(
      x = income %>% factor(levels = unique(housing_burden_epa_reordered$income)),
      y = estimate,
      fill = burden,
    ),
    stat = "identity",
    position = "stack"
  ) +
  labs(
    x = "Household Income",
    y = "Number of Households",
    title = "East Palo Alto housing burden",
    fill = "% of Income spent on housing"
  ) +
  coord_flip() +
  facet_grid(. ~ occupation)

burden_plot_epa

```

The two plots above further illuminated the situation in East Palo Alto. We can see at a glance that the majority of people in the area earn more than $75,000, and the proportion of the population that is housing burdened in that income tier is relatively similar. Still, 

We can also see that there are more renters in total when looking in the lower income tiers. Furthermore, there is a much higher proportion and total number of renters that are housing burdened then owners for these income teirs. Let's explore this intuitive understanding by looking closer at the data of renter-occupied units and housing burden. 

(% of renter-occupied units, and what % of severely burdened households are renter households.)

The OPA hopes to give renters more opportunity to purchase housing and thus stabilize communities. To gain a deeper understanding of the renting and burden situation, lets see how many housing units are renter occupied and then look at what percent of burdened households are renter households. 

First, let's look at the breakdown of housing burden to determine just how much housing burden the renting population of Est Palo Alto faces. This will help us to see how valid the OPA's goal of assisting the renting population is, by showing how much more likely they are to be housing burdened. 

Below is the percentage of housing burdened families there are in East Palo Alto that are renters.

```{r, warning = F}
percent_housing_burden <- housing_burden_epa_reordered %>% 
  group_by(occupation, burden) %>% 
  summarize(estimate = sum(estimate)) %>% 
  filter(burden == "30 percent or more") %>% 
  mutate(
    percentage = estimate / sum(percent_housing_burden$estimate)
  )

percent_housing_burden[2,4]
```

It's clear that most of the housing burdened population for East Palo Alto are renters, with about 66% of the housing burdened being renters. This shows us that the OPA does at least have a valid concern in attempting to relieve housing burden by targeting the renters of the area if there focus is to help those most impacted.

Now, let's look at how many of the housing units are actually renter-occupied, to see how many people the OPA would be trying to impact. Knowing how many renters there are in East Palo Alto to begin with will help us understand if the impact is large enough to be warranted.

Below is the percentage of housing units that are renter-occupied in East Palo Alto. 

```{r, warning = F}
percent_renter_EPA <- housing_burden_epa_reordered %>% 
  group_by(occupation) %>% 
  summarize(estimate = sum(estimate)) %>% 
  mutate(
    percentage = estimate / sum(estimate)
  )

percent_renter_EPA[2,3]
```

The percentage of renter-occupied housing units in East Palo Alto is about 59.3%, which means the majority of housing is occupied by the renters that the OPA is looking to help. 



Parcel Data

```{r}
epa_zoning <- esri2sf("https://services8.arcgis.com/0IC8hha3hgQkmGoa/arcgis/rest/services/EastPaloAltoZoning_2021_WFL1/FeatureServer/1")

smc_exemption <- read_csv("https://datahub.smcgov.org/api/views/pmvw-bs84/rows.csv?accessType=DOWNLOAD")

epa_exemption <- smc_exemption %>% 
  mutate(
    APN = `Parcel Number` %>% 
      str_replace_all("-","")
  ) %>% 
  filter(APN %in% epa_zoning$APN) %>% 
  left_join(epa_zoning) %>% 
  st_as_sf() %>% 
  filter(New_Zone == "R-LD")

leaflet() %>% 
  addProviderTiles(provider = providers$CartoDB.Positron) %>% 
  addPolygons(
    data = epa_exemption %>% 
      filter(`Fiscal Year` == "2018/2019"),
    fillColor = "yellow",
    color = "black",
    weight = 0.5
  ) %>% 
  addPolygons(
    data = epa_exemption %>% 
      filter(Exemption %in% c(5600,7000)),
    fillColor = "blue",
    color = "black",
    weight = 0.5,
    fillOpacity = 1
  )

```


