---
title: "ChrisEscobedo_A3"
author: "Chris"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
---

```{r, include = F}
knitr::opts_chunk$set(echo = F, warnings = F, message = F)
```

```{r}
library(lehdr)
library(tidyverse)
library(tigris)
library(sf)
library(leaflet)
library(censusapi)
library(mapboxapi)

Sys.setenv(CENSUS_KEY="c7ce771b506b13ca6875d12a179606f8087fa381")
```

## Introduction

This analysis will focus on the zip code area of 94804, which is in the South of Richmond. This area is intersting to analyze, as it has much lower median income and is much more diverse than many of the surrounding. This might imply that more of the members of this community will commute to work, as they are not able to afford areas closer to economic opportunities. It also will be interesting to look at the energy usage, given homes and buildings may be older and not as likely to be retrofitted because of the lack of capital for such upgrades. This context might provide valuable insight into how funding and support for upgrades in communities like this and affordable housing upgrades closer to job hubs could make large differences for reaching our GHG and equity goals as a state.

## Vehicle Emissions Analysis

To create the analysis of vehicle emissions, this report uses data from LODES for trip data. It uses data from 2013 to 2019 to calculate commute emissions for the zip code area in the south of Richmond as an origin and as a destination.

```{r, warning = FALSE}
# zctas <- zctas()
# 
# zip <- zctas %>% 
#   filter(GEOID10 == 94804)
# 
# blocks <- blocks("CA")
# 
# zip_blocks <- blocks %>% 
#   st_centroid() %>% 
#   .[zip, ]
# 
# saveRDS(zip_blocks, "zip_blocks.rds")

zip_blocks <- readRDS("zip_blocks.rds")

## compiling OD LODES data from 2013-2019
# full_zip_od <- 2013:2019 %>% 
#   map_dfr(function(year){
#     
#     print(year)
#     
#     temp <- read_csv(paste0("/Volumes/GoogleDrive/Shared drives/SFBI/Data Library/LODES/ca_od_main_JT01_", year, ".csv.gz")) %>% 
#       filter(
#         h_geocode %in% zip_blocks$GEOID10 |
#           w_geocode %in% zip_blocks$GEOID10
#       ) %>% 
#       mutate(year = year)
#     
#     saveRDS(temp, paste0("temp_od_", year, ".rds"))
#     
#     return(temp)
#     
#   })
# 
# saveRDS(full_zip_od, "full_zip_od.rds")

full_zip_od <- readRDS("full_zip_od.rds")

## Removing "negligible trips" - e.g. trips that went from one zip code to the same zip code
full_zip_od_clean <- full_zip_od %>% 
  select(-createdate) %>% 
  filter(!(
    h_geocode %in% zip_blocks$GEOID10 &
      w_geocode %in% zip_blocks$GEOID10
  )) %>% 
  mutate(
    direction = ifelse(
      h_geocode %in% zip_blocks$GEOID10,
      "outbound",
      "inbound"
    )
  )
```

*This analysis assumes that commute data originating and ending in the same census block would be negligible, especially when compared to trips crossing boundaries of census block boundaries. Because this analysis groups trip origin and destination points by census block, the commute data that is within a single census block is disregarded and treated as zero. Of course, this will remove a non-zero number of trips and emissions from the final analysis. The assumption is that this omission will not likely constitute a large enough amount of emissions that the advantages of grouping by the census block to simplify the analysis will outweigh the disadvantages.*



```{r}
full_zip_od_routing <- full_zip_od_clean %>% 
  mutate(
    origin = ifelse(
      direction == "inbound",
      h_geocode,
      w_geocode
    ),
    cbg = origin %>% substr(1,12),
    tract = origin %>% substr(1,11)
  ) %>% 
  filter(!duplicated(cbg))

ca_tracts <- tracts("CA")

zip_od_origin <-
  full_zip_od_routing %>% 
  select(tract) %>% 
  left_join(ca_tracts %>% select(tract = GEOID)) %>% 
  st_as_sf() %>% 
  st_centroid() %>% 
  st_coordinates()

zip_od_destination <-
  zip %>% 
  st_centroid() %>% 
  st_coordinates()
  
zip_od_route <- 
  1:nrow(zip_od_origin) %>%
  map_dfr(function(x){
    
    if(x%%100 == 0) print(x)
    
    tryCatch(
      mb_directions(
        origin = zip_od_origin[x, ],
        destination = zip_od_destination,
        profile = "driving-traffic"
      ) %>% 
        mutate(id = x),
      error = function(e){
        data.frame(id = x)
      }
    )
    
    
  }) %>% 
  st_as_sf()


```





### Finish routing and vehicle information for thursday












