---
title: "A1_ChrisEscobedo"
author: "Chris"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document  
---

```{r, include = F}
knitr::opts_chunk$set(echo = F, warnings = F, message = F)
```

```{r, warning = F}
library(tidyverse)
library(mapview)
library(tigris)
library(sf)
library(leaflet)
library(esri2sf)
library(tidycensus)
library(censusapi)
library(readxl)

Sys.setenv(CENSUS_KEY="c7ce771b506b13ca6875d12a179606f8087fa381")

census_api_key("c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")

acs_vars_2019_5yr <-
  listCensusMetadata(
    name = "2019/acs/acs5/subject",
    type = "variables"
  )
```

## Brief Introduction

This report assumes familiarity with the OPA and its motives for the city of East Palo Alto.

This report will attempt to analyze the current housing situation, particularly for those who the OPA will attempt to impact the most, to gain an understanding of its merits and the concerns of its opposition. To do that, the report will look at the housing burden of East Palo Alto, the changes in home ownership and occupation status, as well as general housing demographics.

#### Housing Burden

For the purpose of this assessment, housing burden has been defined as using by a threshold of 30% of one's household income. Those who spend more than 30% of income on housing are considered to be housing burdened, while those spend less are considered to not be housing burdened. Although this is an arbitrarily chosen cutoff, it is useful to use to get a quick understanding of those who may struggle to pay for housing in the area.

That being said, it is important to note there may be times when this makes the findings less accurate. For extremely high income earners, using more than 30% on housing may not matter to their living situation if they have plenty of disposable income leftover. Furthermore, some families that use 29% of their income on housing may be struggling financially, but will not be counted in this measure of housing burden because of the arbitrary cutoff. 


```{r, warning = F}
neighboring_counties <- 
  counties("CA", cb = T, progress_bar = F) %>% 
  filter(NAME %in% c("San Mateo","Santa Clara"))

ca_places <- 
  places("CA", cb = T, progress_bar = FALSE) 

neighboring_places <- 
  ca_places %>% 
  st_centroid() %>% 
  .[neighboring_counties, ] %>% 
  st_drop_geometry() %>% 
  left_join(ca_places %>% select(GEOID)) %>% 
  st_as_sf() 

EPA_boundary <- 
  places("CA", cb = T, progress_bar = FALSE) %>% 
  filter(NAME == "East Palo Alto")

housing_burden_neighboring_counties <- getCensus(
  name = "acs/acs5/subject",
  vintage = 2019,
  region = "place:*",
  regionin = "state:06",
  vars = paste0("S2503_C0", c(
  "3_025E","3_026E", "3_027E", "3_028E", "3_029E","3_030E", "3_031E", "3_032E", "3_033E", "3_034E", "3_035E", "3_036E", "3_037E", "3_038E", "3_039E", "3_040E", "3_041E", "3_042E", "3_043E", "3_044E", "3_045E", "3_046E", "5_025E","5_026E", "5_027E", "5_028E", "5_029E","5_030E", "5_031E", "5_032E", "5_033E", "5_034E", "5_035E", "5_036E", "5_037E", "5_038E", "5_039E", "5_040E", "5_041E", "5_042E", "5_043E", "5_044E", "5_045E", "5_046E")
  )
) %>% 
  filter(place %in% neighboring_places$PLACEFP) %>%
  select(!c(state,place) & !ends_with(c("EA","MA","M"))) %>%
  pivot_longer(
    ends_with("E"),
    names_to = "name",
    values_to = "estimate"
  ) %>%
  left_join(
    acs_vars_2019_5yr %>% 
      select(name, label)
    ) %>% 
  select(-name) %>% 
  separate(
    label,
    into = c(NA,"occupation",NA, NA,"income", "burden"),
    sep = "!!"
  ) %>% 
  filter(!is.na(burden)) %>% 
  mutate(
    burden = if_else(burden == "20 to 29 percent" | burden == "Less than 20 percent", 
              "less than 30 percent", 
              burden)
  ) %>% 
  group_by(income, occupation, burden) %>% 
  summarize(estimate = sum(estimate))

housing_burden_epa <- getCensus(
  name = "acs/acs5/subject",
  vintage = 2019,
  region = "place:20956",
  regionin = "state:06",
  vars = paste0("S2503_C0", c(
  "3_025E","3_026E", "3_027E", "3_028E", "3_029E","3_030E", "3_031E", "3_032E", "3_033E", "3_034E", "3_035E", "3_036E", "3_037E", "3_038E", "3_039E", "3_040E", "3_041E", "3_042E", "3_043E", "3_044E", "3_045E", "3_046E", "5_025E","5_026E", "5_027E", "5_028E", "5_029E","5_030E", "5_031E", "5_032E", "5_033E", "5_034E", "5_035E", "5_036E", "5_037E", "5_038E", "5_039E", "5_040E", "5_041E", "5_042E", "5_043E", "5_044E", "5_045E", "5_046E")
  )
) %>% 
  filter(place %in% neighboring_places$PLACEFP) %>%
  select(!c(state,place) & !ends_with(c("EA","MA","M"))) %>%
  pivot_longer(
    ends_with("E"),
    names_to = "name",
    values_to = "estimate"
  ) %>%
  left_join(
    acs_vars_2019_5yr %>% 
      select(name, label)
    ) %>% 
  select(-name) %>% 
  separate(
    label,
    into = c(NA,"occupation",NA, NA,"income", "burden"),
    sep = "!!"
  ) %>% 
  filter(!is.na(burden)) %>% 
  mutate(
    burden = if_else(burden == "20 to 29 percent" | burden == "Less than 20 percent", 
              "less than 30 percent", 
              burden),
    occupation = if_else(occupation == "Owner-occupied housing units", 
              "Owner", 
              "Renter"),
  ) %>% 
  group_by(income, occupation, burden) %>% 
  summarize(estimate = sum(estimate))

housing_burden_epa_reordered <- housing_burden_epa %>% 
  filter(income == "Less than $20,000") %>% 
  rbind(housing_burden_epa) %>% 
  .[-c(21, 22, 23, 24), ]



```

### Housing Burden in East Palo Alto 

Below are two plots that represent general housing burden in East Palo Alto. The plots are separated by occupational status as either a "renter" or "owner" of the house. Since the OPA is interested in assisting the renter-occupied families who are renting housing in building wealth and preventing their displacement, it is helpful to gain this general understanding of the renting and housing burden landscape.

```{r, warning = F}
burden_plot_epa <- housing_burden_epa_reordered %>% 
  ggplot() +
  geom_bar(
    aes(
      x = income %>% factor(levels = unique(housing_burden_epa_reordered$income)),
      y = estimate,
      fill = burden,
    ),
    stat = "identity",
    position = "stack"
  ) +
  labs(
    x = "Household Income",
    y = "Number of Households",
    title = "East Palo Alto housing burden",
    fill = "% of Income spent on housing"
  ) +
  coord_flip() +
  facet_grid(. ~ occupation)

burden_plot_epa

```

The two plots above further illuminated the situation in East Palo Alto. We can see at a glance that the majority of people in the area earn more than $75,000, and the proportion of the population that is housing burdened in that income tier is relatively similar. Still, 

We can also see that there are more renters in total when looking in the lower income tiers. Furthermore, there is a much higher proportion and total number of renters that are housing burdened then owners for these income teirs. Let's explore this intuitive understanding by looking closer at the data of renter-occupied units and housing burden. 

(% of renter-occupied units, and what % of severely burdened households are renter households.)

The OPA hopes to give renters more opportunity to purchase housing and thus stabilize communities. To gain a deeper understanding of the renting and burden situation, lets see how many housing units are renter occupied and then look at what percent of burdened households are renter households. 

First, let's look at the breakdown of housing burden to determine just how much housing burden the renting population of Est Palo Alto faces. This will help us to see how valid the OPA's goal of assisting the renting population is, by showing how much more likely they are to be housing burdened. 

Below is the percentage of housing burdened families there are in East Palo Alto that are renters.

```{r, warning = F}
percent_housing_burden <- housing_burden_epa_reordered %>% 
  group_by(occupation, burden) %>% 
  summarize(estimate = sum(estimate)) %>% 
  filter(burden == "30 percent or more") %>% 
  group_by(occupation) %>% 
  summarize(estimate = sum(estimate)) %>% 
  mutate(
    percentage = estimate / sum(estimate)
  )

percent_housing_burden[2,3]
```

It's clear that most of the housing burdened population for East Palo Alto are renters, with about 66% of the housing burdened being renters. This shows us that the OPA does at least have a valid concern in attempting to relieve housing burden by targeting the renters of the area if there focus is to help those most impacted.

Now, let's look at how many of the housing units are actually renter-occupied, to see how many people the OPA would be trying to impact. Knowing how many renters there are in East Palo Alto to begin with will help us understand if the impact is large enough to be warranted.

Below is the percentage of housing units that are renter-occupied in East Palo Alto. 

```{r, warning = F}
percent_renter_EPA <- housing_burden_epa_reordered %>% 
  group_by(occupation) %>% 
  summarize(estimate = sum(estimate)) %>% 
  mutate(
    percentage = estimate / sum(estimate)
  )

percent_renter_EPA[2,3]
```

The percentage of renter-occupied housing units in East Palo Alto is about 59.3%, which means the majority of housing is occupied by the renters. Thus, since the OPA would be looking to help renters, it would be helping the majority of the families living in East Palo Alto.

Given that information, we have a better understanding of the housing situation in Est Palo Alto. Most of the houses are renter owned. Most of the housing burden falls upon those renters. So there may be merit to the claim of OPA supporters that renters are more housing burdened and that they should be helped first.

Still, we can look closer at parcel level data in East Palo Alto to uncover more patterns in housing in the area. Specifically, we may be interested in how home ownership is changing. The OPA looks to increase home ownership of community members, those living in the area, to stabilize prices and community. By looking at the change in home ownership with the occupation status, we can see what is currently happening with home ownership. 

#### Using Parcel Data to uncover home ownership trends

To get started, lets look at the landscape of East Palo Alto in relation to owner and renter occupation. Those who oppose the OPA worry about the additional eligibility of "single family homes", whereas other cities have foused on multifamily units. This understanding of parcel data will help us to understand the makeup of those single-family homes and see what changes have occurred that might warrant external support in the form on an OPA or not.

The map below uses the report of a tax exemption for owner occupied houses to determine the occupation status of single family homes. 

*Although this is helpful, it may not be fully accurate, as there may be owners who forget to apply for the exemption. There also may be buildings that have a zero value for the exemption but are not currently used for residential purposes, and thus the houses that do not file for the exemption are not necessarily renter-occupied, as the map below assumes. 

Furthermore, the map uses parcels that are zoned for "low density residential" (R-LD), which should get us to single-family properties in the city, but of course, zoning is not actual property use and there may be parcels that are not single family homes.*

```{r}
epa_zoning <- esri2sf("https://services8.arcgis.com/0IC8hha3hgQkmGoa/arcgis/rest/services/EastPaloAltoZoning_2021_WFL1/FeatureServer/1")

smc_exemption <- read_csv("https://datahub.smcgov.org/api/views/pmvw-bs84/rows.csv?accessType=DOWNLOAD")

epa_exemption <- smc_exemption %>% 
  mutate(
    APN = `Parcel Number` %>% 
      str_replace_all("-","")
  ) %>% 
  filter(APN %in% epa_zoning$APN) %>% 
  left_join(epa_zoning) %>% 
  st_as_sf() %>% 
  filter(New_Zone == "R-LD")

leaflet() %>% 
  addProviderTiles(provider = providers$CartoDB.Positron) %>% 
  addPolygons(
    data = epa_exemption %>% 
      filter(`Fiscal Year` == "2018/2019"),
    fillColor = "yellow",
    color = "black",
    weight = 0.5,
    label = ~
      "renter-occupied",
    highlightOptions = highlightOptions(
      weight = 2,
      opacity = 1
    )
  ) %>% 
  addPolygons(
    data = epa_exemption %>% 
      filter(Exemption %in% c(5600,7000)),
    fillColor = "blue",
    color = "black",
    weight = 0.5,
    fillOpacity = 1,
    label = ~
      "owner-occupied",
    highlightOptions = highlightOptions(
      weight = 2,
      opacity = 1
    )
  ) %>% 
  addLegend(
    data = epa_exemption,
    colors =c("yellow", "blue"),
    labels= c("renter-occupied", "owner-occupied"),
    title = "Occupation status of parcel"
  )

```

From this map, we can see that there is a fairly heterogeneous mix of renter and owner occupation of single family homes in East Palo Alto, with mixed geographical diversity and fairly even distribution. 

Now let's look deeper at the changes that have occured to see if any of these groups have become more prevelant. Specifically, we want to see if there is an increasing number of renter-occupied parcels to determine if the parcels are vulnerable to being bought by those outside of the community and rented at higher prices at the expense of the community members. We won't be able to see changes in rent prices, but we can look at the change in occupation status to get a sense of what might be happening. 



#### Conclusion

Going through this analysis, it seems clear to me that the most burdened popeulation in East Palo Alto are the renters. Furthermore, the increase in homes occupied by renters might indicate a lack of community wealth-building and stability for the long time community members, as they are out-competed by those investment interests who purchase properties to rent them out.

The opposition to the OPA believes this may harm single family home owners, but given the exemption to owners who occupy the house, I believe that impact will be fairly small. Furthermore, it is clear that the OPA will be helping a much larger group and a much more burdened group in getting a better chance to increase home ownership levels. 

For these reasons I am in support of the OPA as I see it. Of course there are other complexities that I am not able to analyze through this report, such as what financial costs there may be to those who wish to invest in the community but are now disincentivized to do so. I believe that the good this may do outweighs the potential negatives, but remain open to learning more about it. 










