---
title: "ChrisEscobedo_A5"
author: "Chris"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
---

```{r, include = F}
knitr::opts_chunk$set(echo = F, warnings = F, message = F)
```

```{r, include = F}}
library(tidyverse)
library(sf)
library(leaflet)
library(mapview)
library(corrplot)
library(gtools)
library(car)
library(tigris)
library(censusapi)
library(readxl)

Sys.setenv(CENSUS_KEY="c7ce771b506b13ca6875d12a179606f8087fa381")

acs_vars_2019_5yr <-
  listCensusMetadata(
    name = "2019/acs/acs5",
    type = "variables"
  )
```


```{r}
ces4 <- read_excel("calenviroscreen40resultsdatadictionary_F_2021.xlsx")

ca_tracts <- tracts("CA")

ces4_map <- ces4 %>% 
  left_join(
    ca_tracts %>% 
      transmute(GEOID = GEOID %>% as.numeric()),
    by = c("Census Tract" = "GEOID") 
  ) %>% 
  st_as_sf()

```
## Asthma and PM2.5 Concentration Maps

The following maps show the concentration of asthma and PM2.5 prevalence. The data used to create the maps comes from CalEnviroScreen 4.0, released on October 20, 2021.

#### Asthma Data
The data for the map showing the prevalence of Asthma is the rate of Emergency Department visits for asthma per 10,000 people per year, averaged over 2015-2017. Thus, the data is only up to date until 2017 and may reflect numbers that are no longer accurate. 

```{r, warning = F}
Asthma_pal <- colorNumeric(
  palette = "YlOrRd",
  domain = ces4_map$Asthma
)

leaflet() %>%
  addTiles() %>% 
  addPolygons(
    data =ces4_map,
    fillColor = ~Asthma_pal(Asthma),
    color = "white",
    opacity = 0.5,
    fillOpacity = 0.8,
    weight = 1,
    label = ~paste0(
      round(Asthma), 
      " ED visits for Asthma per year per 10,000"
    ),
    highlightOptions = highlightOptions(
      weight = 2,
      opacity = 1
    )
  ) %>% 
  addLegend(
    data = ces4_map,
    pal = Asthma_pal,
    values = ~Asthma,
    title = "ED visits for Asthma<br>per year per 10,000,<br>averaged over 2015-2017"
  )
```
#### PM2.5 Data
The PM2.5 map contains data on the fine particle pollution throughout the state. The data represents the annual mean concentration of PM2.5 (the weighted average of measured monitor concentrations and satellite observations, µg/m3), over three years, from 2015 to 2017. Thus, the data may also only be accurate up until 2017 and may be outdated. 

```{r}
PM_pal <- colorNumeric(
  palette = "Purples",
  domain = ces4_map$PM2.5
)

leaflet() %>%
  addTiles() %>% 
  addPolygons(
    data =ces4_map,
    fillColor = ~PM_pal(PM2.5),
    color = "white",
    opacity = 0.5,
    fillOpacity = 0.8,
    weight = .5,
    label = ~paste0(
      round(PM2.5), 
      " ED visits for Asthma per year per 10,000"
    ),
    highlightOptions = highlightOptions(
      weight = 2,
      opacity = 1
    )
  ) %>% 
  addLegend(
    data = ces4_map,
    pal = PM_pal,
    values = ~PM2.5,
    title = "ED visits for Asthma<br>per year per 10,000,<br>averaged over 2015-2017"
  )
```

## Asthma and PM2.5 Scatter Plot

```{r}

```











